Towards a Pandemic-Scale Ancestral Recombination Graph of SARS-CoV-2

### 2023-05-02 paper.tex
Fisrt pass at recombinants

\begin{itemize}
\item A potential source of error in sc2ts ARGs comes from incorrectly reported collection dates of the samples, which can cause sample genomes to be attached to the wrong parts of the ARGs. Thus far, we have attempted to correct the most offending errors in the ARGs by filtering out "time travellers", but the sampling time can still be wrong by days or tens of days. ***describe possible approaches to correct this?***
\item To avoid time travellers, we used a maximum submission delay, but this will systematically bias against submissions from countries with greater submission delays (Kalia et al.). An area of future work is to add back the excluded sequences to a built ARG to assess which ones are likely true time travellers.
\item We have used a single mismatch value in all the HMM runs, set so that having three mutations is as likely as having a single recombination. This parameter choice is arbitrary, and higher values may reduce the number of origins for those Pango lineages that were inferred to descend from more than one recombinant ancestor. In addition, because both the mutation rate and recombination rate varies across the SARS-CoV-2 genome, it may be helpful to allow the mismatch value to vary along the genome.
\item By treating mutations equally in the current HMM implementation, we implicitly assume that the rates of different types of mutations are identical. However, as reported previously \citep{Yi2021-sc}, there is a high degree of mutational asymmetry in SARS-CoV-2. To account for mutational biases, more sophisticated nucleotide substitution models (e.g., allowing for different rates of transition and transversion) can be incorporated into HMM likelihood calculations.
\item Our current masking strategy handles indels as missing data. Doing this can remove information helpful to detect recombinants. For example, sequences of the Pango recombinant XP were not identified as recombinants, because XP’s characteristic multibase deletion was masked.
\item The "lack of memory" in the HMM can result in  recombinants with two or more breakpoints to have different parents for each recombined portion of the genome, when it is sometimes equally parsimonious to match against the same parent in separate regions. The copying process in the HMM currently does not preferentially "switch back" to a previoulsy matched parent, but incorporating this would be a valuable addition.
\item Some recombinants may reasonably have three or more parents, indicating multiple "stacked" recombination events. We have not investigated this in the current analysis, but with further sc2ts development, there is potential to analyse these putative recombinants and trace their evolutionary origins through the ARG.
\item In the present implementation, HMM traceback is performed once, yielding a single maximum-likelihood genome copying path. A stochastic traceback algorithm (e.g., (Rasmussen et al. 2014)) can be implemented in order to quantify uncertainty about putative parent node(s) and switch positions. These estimates of uncertainty may be used as support values for branches analogous to bootstrap values produced by standard phylogenetic methods.
\item Daily trees are built using the UPGMA algorithm. This entails an extra step to ``un-resolve'' daily trees in parts of the trees with no phylogenetic resolution. Using a maximum-parsimony algorithm to build daily trees would leave these "soft polytomies" in, removing the need for the un-resolving step.
\end{itemize}

@@

[TODO ADD IN HMM RERUN STUFF]

@@

% \subsubsection{Breakpoint intervals and HMM consistency} When recombinant
% ancestry for a given genome is inferred by the LS HMM, a breakpoint is
% identified as the last possible position where the copying path must switch.
% Thus, the breakpoints used in the ARG represent the rightmost end of the
% interval, within which the recombination event could have occurred. To obtain
% the leftmost end of this interval, we rerun the HMM using mirrored coordinates
% (i.e., from the 3’ end to the 5’ end).

>>>>

\begin{itemize}
\item A potential source of error in sc2ts ARGs comes from incorrectly reported collection dates of the samples, which can cause sample genomes to be attached to the wrong parts of the ARGs. Thus far, we have attempted to correct the most offending errors in the ARGs by filtering out "time travellers", but the sampling time can still be wrong by days or tens of days. ***describe possible approaches to correct this?***
\item To avoid time travellers, we used a maximum submission delay, but this will systematically bias against submissions from countries with greater submission delays (Kalia et al.). An area of future work is to add back the excluded sequences to a built ARG to assess which ones are likely true time travellers.
\item We have used a single mismatch value in all the HMM runs, set so that having three mutations is as likely as having a single recombination. This parameter choice is arbitrary, and higher values may reduce the number of origins for those Pango lineages that were inferred to descend from more than one recombinant ancestor. In addition, because both the mutation rate and recombination rate varies across the SARS-CoV-2 genome, it may be helpful to allow the mismatch value to vary along the genome.
\item By treating mutations equally in the current HMM implementation, we implicitly assume that the rates of different types of mutations are identical. However, as reported previously \citep{Yi2021-sc}, there is a high degree of mutational asymmetry in SARS-CoV-2. To account for mutational biases, more sophisticated nucleotide substitution models (e.g., allowing for different rates of transition and transversion) can be incorporated into HMM likelihood calculations.
\item Our current masking strategy handles indels as missing data. Doing this can remove information helpful to detect recombinants. For example, sequences of the Pango recombinant XP were not identified as recombinants, because XP’s characteristic multibase deletion was masked.
\item Daily trees are built using the UPGMA algorithm. This entails an extra step to ``un-resolve'' daily trees in parts of the trees with no phylogenetic resolution. Using a maximum-parsimony algorithm to build daily trees would leave these "soft polytomies" in, removing the need for the un-resolving step.
\end{itemize}

@@

When the Viterbi algorithm implementation used by the LS HMM infers
recombinant ancestry for a given genome, the point at which inheritance
switches from one parent node to another is the last possible
position.
Thus, the breakpoints used in the ARG represent the rightmost end of the
interval, within which the recombination event could have occurred. To obtain
the leftmost end of this interval, we rerun the HMM using mirrored coordinates
(i.e., from the 3’ end to the 5’ end) and store the information as
metadata on the recombination node.
These left- and rightmost positions
are the recombination intervals reported throughout.
Ultimately all the recombination intervals mean is that the parent sequences
are identical within that interval, and there is therefore no information
about where the recombination occured. There are many different processes
that can contribute to the size of these intervals, and therefore
understanding how these processes interact to produce wider or narrower
intervals is an important avenue for future work.

The LS HMM machinery, and the interpretation of inferred recombinant paths and
breakpoint intervals is a central part of \texttt{sc2ts}, and there are many
ways to extend and improve. For example, the current parameterisation of using
a single ``mismatch ratio'' is very simplistic (see section XXX), and likely
results in a very flat likelihood space where many sequences have equal
likelihood of being chosen. More sophisticated parameterisations where we
condition on known per-site mutation rates (and perhaps per-base subsitution
rates) may lead to better results, although perhaps at the cost of some
reduction in interpretability. Post-processing the match results to produce
more parsimonious breakpoints may also be a worthwhile avenue for development.
In particular, we may choose to insert breakpoints for the ARG that are choosed
from within the possible interval, rather than the current approach of taking
the right-most value. Cases where we have more than two parents may either be
the ``stacking'' of multiple recombination events, or instances where
the HMM has chosen to switch to a third sequence rather than back to
the original parent (where this is equally parsimonious). Many putative
recombination events, however, will represent poor quality data, where
a recombinant copying path happens to be a more likely explanation
of a highly divergent sequence.
A thorough analysis of the behaviour of the LS model in the context
of a pandemic-scale ARG may lead to significant improvements in our
ability identify recombinants and to filter poor quality data.

@@
